---
title: "Schmalfeld_606Fall2021_Final_Project_Happiness_correlation_with wellness"
author: "Mark Schmalfeld"
date: "10/31/2021"
output:
  html_document: default
  pdf_document: 
    latex_engine: xelatex
---
---
title: DATA 607 Fall 2021 Final Project-Energy Forecast
author: Mark Schmalfeld
---


```{r echo=FALSE}
library(tidyverse)
library(openintro)
library(RCurl)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Energy and Carbon Emissions data upload

#### Use the US Energy Information Administration (EIA) at www.eia.gov.international to collect following

#### Collect data on a world and regional level


### Pair down the time series data for the review at the global level

1) Population dataset
2) GDP
3) Carbon Emissions data
4) Energy production and demand 
5) Energy use per capita
6) Energy use per GDP

Subset the data and rename to clean up the variable names.
Create a data frame with import global varialbes in time series ( aprpox 40 year of annual data)
Look at the glimpse of the dataset to see basic structure

```{r}

URL<-"https://raw.githubusercontent.com/schmalmr/607Finalproject/main/EIAdata607FinalProject_Energy.csv"

df<-read.csv(URL)


dfglobal<-select(df,Year, World_Consumption_quadBTU, World_GDP_USDB, World_consume_Coa._quadBTU, World_consume_Natural.gas_quadBTU,World_consume_Petroleum_quadBTU, World_consume_Nuclear..renewables_quadBTU,World_pop,World_MMBtu_person, Fossil.fuel_plus_landuse_emissions_GtCO2,Land_use_emissions_GtCO2, Fossil_fuel_industry_emissions_GtCO2, World_1000BTU_GDPBUSD )


dfglobal<-rename(dfglobal, gobal_zero_CO2_energy= World_consume_Nuclear..renewables_quadBTU)
dfglobal<-rename(dfglobal, global_population =World_pop)
dfglobal<-rename(dfglobal, global_CO2_fuels=Fossil_fuel_industry_emissions_GtCO2)
dfglobal<-rename(dfglobal, MBTU_per_GPD_Billions=World_1000BTU_GDPBUSD)
dfglobal<-rename(dfglobal, MBTU_per_person=World_MMBtu_person)
dfglobal<-rename(dfglobal, global_energy_use_quad_BTU= World_Consumption_quadBTU)
dfglobal<-rename(dfglobal, global_Coal_use_quad_BTU=World_consume_Coa._quadBTU)
dfglobal<-rename(dfglobal, global_NGAS_use_quad_BTU=World_consume_Natural.gas_quadBTU)
dfglobal<-rename(dfglobal, global_Oil_use_quad_BTU=World_consume_Petroleum_quadBTU)
dfglobal<-rename(dfglobal, global_fuelandlandland_use_GtCO2=Fossil.fuel_plus_landuse_emissions_GtCO2)
dfglobal<-rename(dfglobal, global_land_use_GtCO2=Land_use_emissions_GtCO2)

glimpse (dfglobal)

```

```{r create plots to view datasets}

ggplot(data=df, aes(x=Year, y=World_Consumption_quadBTU))+geom_point(size=4)+geom_smooth(method=lm)+ggtitle("Energy consumption (quad) vs time")

ggplot(data=df,aes(x=World_pop, y=World_Consumption_quadBTU))+geom_point(size=5)+geom_smooth(method=lm)+ggtitle("Energy consumption (quad) vs global population")

ggplot(data=df,aes(x=World_pop, y=Fossil_fuel_industry_emissions_GtCO2))+geom_point(size=4)+geom_smooth(method=lm)+ggtitle("Fossile fuel carbon emission (GtCo2) vs world population")

ggplot(data=df,aes(x=World_Consumption_quadBTU, y=Fossil_fuel_industry_emissions_GtCO2))+geom_point(size=4)+geom_smooth(method=lm)+ggtitle("Energy consumption (quad) vs fossile fuel emissions (GtCO2)")

ggplot(data=df,aes(x=World_Consumption_quadBTU, y=World_GDP_USDB.1))+geom_point(size=4)+geom_smooth(method=lm)+ggtitle("Energy consumption (quad) vs GDP (billions USD)")

ggplot(data=df,aes(x=Fossil_fuel_industry_emissions_GtCO2, y=World_GDP_USDB.1))+geom_point(size=4)+geom_smooth(method=lm) + ggtitle("Fossil fuel emissions (GtCO2) vs GPD (US $ Billions)")  
       
```


## Evaluate linear model of GPD vs year

```{r linear model evaluation GPD at given time}


 m_gdp <- lm(World_GDP_USDB ~ Year, data = dfglobal)
summary(m_gdp)
par(mfrow = c(2,2))
plot(m_gdp)

```

## Evaluate linear model of global population over time 

```{r linear model evaluation population at given year}


 m_population2 <- lm(global_population ~ Year, data = dfglobal)
summary(m_population2)
par(mfrow = c(2,2))
plot(m_population2)

```

##  Evalue Energy consumption total 

### Model is good with the GPD and the Population.
#### Population has a p value of 0.06 (just a bit higher than desired)
#### Population retained as a variable in the model 

```{r linear model evaluation energy at with GPD and Population}


 m_energy <- lm(global_energy_use_quad_BTU ~World_GDP_USDB+global_population, data = dfglobal)
summary(m_energy)
par(mfrow = c(2,2))
plot(m_energy)

```

## Evaluate the Energy consumption not generating CO2 

### Correlation is most strong with the Global GDP 
### GDP calculated from model above can be used to estimate model Energy 
consumption that is not generating CO2.

### Assumption for ease of model: Nuclear, renewables and water
are net zero CO2 energy sources in this model 
All CO2 emissions are then attributed to fuel use.


```{r linear model evaluation energy zero CO2 over time}


 m_energyzero <- lm(gobal_zero_CO2_energy ~ World_GDP_USDB, data = dfglobal)
summary(m_energyzero)
par(mfrow = c(2,2))
plot(m_energyzero)

```
## Model of Carbon Emissions from Fuel over time

## Key variables for the model are the global energy use, global zero CO2 energy 
## and the global population

#### Note:  For simplification the use of carbon capture technology is ignored
and it is assumed later as a simple application of the technology
to remove carbon from the atmosphere. It is actually a very challenging and
difficult problem to solve economically and with efficent use of energy.


```{r linear model evaluation CO2 over time}


 m_co2 <- lm(global_CO2_fuels ~ global_energy_use_quad_BTU+gobal_zero_CO2_energy+global_population, data = dfglobal)
summary(m_co2)
par(mfrow = c(2,2))
plot(m_co2)

```
## Modeling the future 

### The past 40 years provide a dataset used to establish a frame set of models 

1) Models are noted to have some weaknesses in the models; however all of the
models are very good at explaining the past.  The R-squared values for 
the series of models is above 0.95 for all models (amazingly strong)
2) There are some indications of the residuals have a few outlyers in the models
- particually noted in the Carbon Emissions model (some trends in each of models)
3) The Q-Q plots are pretty good for models except the population model over time
which has more drift than others.  Others do have some trends at the start
and the tails of the plots.
4) Residuals are show good random variability for models except the 
     - population model seems to have poor randomness - with a distinct line
     - GPD have drift in residuals up and down across the model
     
5) Major assumption for next step is to proceed to extrapolate the future
with the current model.  
 - Extrapolation out until 2050
 - Note the use of an extrapolation is not recommended and error expected
 - Extrapolation judged a risk to consider based the very strong R-squared for
 each of the individual models.
 
 
6) Create an incremental estimate to demonstrate the magnitude of the incremental 
renewable or zero carbon generating energy development is needed.  
Placeholder globally without specific constraints on population or GDP

7) The appraoch is to show how much more is needed to just hold Carbon emissions
flat or near constant.  The later adjustments show how difficult or significant
the energy transition needed is to achieve reductions on Carbon emissions.


```{r model use}

year <- c(1980,1990,2000,2010,2019, 2020, 2021,2022,2023,2024,2025,2026,2027,2028,2029,2030,2035,2040,2045,2050,2075,2100)

df_model <- data.frame(year)

df_model<- df_model %>%mutate(GDP=(-5033000+2552*year),.after=year)
df_model<- df_model %>%mutate(population=(-170700000+88370*year),.after=GDP)
df_model<- df_model %>%mutate(energy_total=(273.5+0.004*(GDP)),.after=population)
df_model<- df_model %>%mutate(energy_zerocarbon=(19.02+0.0005592*(GDP)),.after=energy_total)
df_model<- df_model %>%mutate(globalcarbonfuels=(-1133000000+6.91E7*energy_total-100900000*energy_zerocarbon+754.7*population),.after=population)

df_model<- df_model %>%mutate(energywithcarbon=(energy_total-energy_zerocarbon),.after=energy_zerocarbon)
df_model<- df_model %>%mutate(carbon_per_energy=(globalcarbonfuels/(energy_total-energy_zerocarbon)),.after=energywithcarbon)

df_model$incremental_new_zero_carbonenergy <- c(0,0,0,0,0,20,35,55,75,95,115,150,170,180,190,195,250,500,650,800,900,1100)
df_model<-df_model %>% mutate(energy_new_zerocarbon=energy_zerocarbon+incremental_new_zero_carbonenergy)
df_model<-df_model %>% mutate(energy_new_withcarbon=(energy_total-(energy_new_zerocarbon)))
df_model<-df_model %>% mutate(new_global_carbonemissionsfuel=(globalcarbonfuels+72000000*(energy_zerocarbon-energy_new_zerocarbon)))






```






```{r plots for modeling, echo=FALSE}

ggplot(data=df_model, aes(x=year, y=energy_zerocarbon))+geom_col(color="red")+ggtitle("Zero Carbon Energy Quads vs time baseline - no accelerated reduction in carbon use")

ggplot(data=df_model,aes(x=year, y=energy_total,size=globalcarbonfuels))+geom_point()+ggtitle("Energy demand forecast and associated carbon emission trend in progress")

ggplot(data=df_model, aes(x=energy_total,y=globalcarbonfuels, size=energy_zerocarbon))+geom_point(size=5)+ggtitle("Energy demand with carbon emissions forecast based on past 40 yr trend")

ggplot(data=df_model, aes(x=year, y=energy_new_zerocarbon))+geom_col(color="green")+ggtitle("Zero Carbon Energy quads with accelerated carbon emissions reduction")

ggplot(data=df_model,aes(x=year, y=energy_total,size=new_global_carbonemissionsfuel))+geom_point()+ggtitle("Energy demand forecast and associated carbon emission trend in progress")

ggplot(data=df_model, aes(x=energy_total,y=new_global_carbonemissionsfuel,size=energy_new_zerocarbon))+geom_point(size=5)+ggtitle("Energy demand with carbon emissions forecast based on past 40 yr trend")

ggplot(data=df_model,aes(x=population,y=GDP,size=energy_total))+geom_point()+ggtitle("Population vs GDP with bubble size based on total energy demand in quad")



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

##  Illustration of the impact energy shift away from fossil fuels

Assumptions for solar:
1)  One square meter of solar cells generates 400 Kw hr per year and efficiency is 70% sunshine at 8 hr per day.
2)  Therefore:  1 Quad= 10^15 BTU of energy or will require 755 Km2 of solar panels
3)  Also 1 quad is approx equavlent to 183 million barrels of oil, 38.5 Million tons of coal or 980 billion cubic feet of natural gas.
4) How do we conceptualize the scale of the task? 
   a) Two refinery locations at 250,000 BPD of oil processing is 1 Quad in a year (combined largest complex in the world that would exist today)
   b) Or appox 500 quad of energy would require a land area the size of Japan, Germany or a bit smaller than Spain.
      or approx 100 quad of energy would require a land area the size of Pannama or Ireland.
  c) or 1.2 Million wind turbines to generate 100 quads of energy which could fit into an area the scale of Delaware.




```{r measurable impact of the reduction targets }


df_model<- df_model %>%mutate (Solar_total_energy_kmsquared=(energy_total*755),.after=year)
df_model<- df_model %>% mutate (Solar_totalenergy_num_Irelands=(Solar_total_energy_kmsquared/75500))

df_model<-df_model %>% mutate(Solar_renewable_energy_kmsquared=(energy_new_zerocarbon*755))
df_model<-df_model %>% mutate(Solar_renewable_num_Irelands=(Solar_renewable_energy_kmsquared/75500))

df_model<-df_model %>% mutate(wind_turbines_totalenergy=(energy_total/100*1200000))
df_model <-df_model %>% mutate(wind_turbines_totalenergy_num_delawares=(energy_total/100))

```





## Plot relative measures of tasks scale to reduce CO2 Emissions with renewables 

Renewable energy as solar or as wind turbines to supply total energy demand 

Equivalent areas estimated to provide benchmark of the scale of the requirements



```{r plot scale examples of solar and wind turbine replacement of energy, echo=FALSE }

ggplot(data=df_model, aes(x=year, y=Solar_totalenergy_num_Irelands, size=energy_total))+geom_point()+ggtitle(" Solar: #Ireland sized areas to meet the global energy demand (energy Quads")

ggplot(data=df_model, aes(x=year, y=wind_turbines_totalenergy_num_delawares, size=energy_total))+geom_point()+ggtitle("Wind Turb # Delaware sized areas required for global energy demand (energy Quads")

```



##  Conclusions

#### ~40 years of data from EIA provides good linear modules to represent the trend in energy demand, carbon emissions from fuel as it relates to GDP, Population  (great R-squared correlations with multiple equation model to use in forecasting) 
#### Model useful to drive understanding of task at hand to create a lower carbon future through forecasting into the future.

## Future work/ improvements:
#### Evaluate model on regional basis and further utilize regional fuel mix.
#### Evaluate additional information on specific shifts in middle class, poverty, and urbanization related to energy use.
#### Evaluate more of the renewable mix sensitivity and include further economics (costs, benefits and non-linear growth potential)
### Evaluate stronger means of developing forecast forward - extrapolation generally not recommended however based on strength of R-squared and p-values and the statistics is seems a good means of improving understanding.




## References

1. US Energy Information Administration (EIA) at www.eia.gov.international

2. 2020 BP World Energy Outlook at  www.bp.com

3. Jefferson W. Tester, et all, (2005) “Sustainable Energy Choosing Among Options” (MIT press)

4. Hadley Wickham, Garret Grolemund.(2017) “R for Data Science” (OReilly)

5. Winston Chang (2019), “R Graphics Cookbook” (OReilly)
